Dark Control:\\ A Unified Account of Default Mode Function\\ by Control Theory and Reinforcement Learning

### 2017-04-06 paper/paper.tex
more fixups
sentinel hypothesis 3.0

\subsection{Putting everything together}
The DMN is today known to consistently increase in neural
  activity when humans engage in goal-directed behavior that are detached from
  current sensory environment \citep{kenet2003spontaneously, fiser2004small}
  and it was proposed to be situated at the top of the functional network hierarchies
  \citep{carhart2010default, margulies2016situating}.
  Its involvement in thinking about the past,
  the future, and hypothetical possibilities ties in with the implicit computation of
  action and state cascades as a function of what happened in the past.
  A policy represents the repertoire of possible actions
    on the world given a current state. It encodes the probabilites of
    choosing actions to be executed in a certain situation.
% Interactions between external and internal sources of
% information are probably commonplace during perception \citep{gershman2010learning}
\begin{figure}[!h]
  \includegraphics[width=.9\linewidth]{neurobiological_and_rl_overview_DMN.pdf}
  \caption{Neurobiological implementation of RL by the DMN...}
  \label{fig:rl_process_chart}
\end{figure}

The DMN may subserve
  constant exploration of possible future actions and their
  cumulative reward outcomes. Implicit computation of future choices
  provides an explanation for the
  evolutionary emergence and practical usefulness of mind-wandering
  in humans.

@@

\subsection{Sentinel hypothesis}
Processing self-relevant information was perhaps the first
cognitive account that was proposed for the DMN \citep{gusnard2001medial}.
Since then,
many investigators have speculated that neural activity in the DMN
may reflect the brain's continuous tracking of
relevance in the environment
as an advantageous evolutionary adaptation.
According to this cognitive account, the human brain's baseline realizes
a ``radar'' function to
detect subjectively salient objects and unexpected events in the environment.
Propositions of such a sentinel function have however seldom detailed
the mechanisms of
how attention and memory resources are exactly reallocated when
encountering a salient environmental stimulus.
In our formal account,
cascades of candidate actions
are recursively explored by human agents. Conversely,
certain branches of candidate action trajectories
are detected to be less worthy to get mentally explored.
This mechanism, expressed by the Bellman equation,
directly implies stratifications over attention and working memory
to relevant objects and events in the environment.
%
Further,
our account provides an explanation for the known implication
of the DMN
in certain goal-directed experimental tasks and
in the idlying task-unconstrained cognitive sets.
Both cognitive sets may entail DMN recruitment
if environmental relevance is processed, manipulated, and used for prediction.
During tasks,
the policy and value estimates are updated to optimize short-term action.
At rest, these parameter updates may
improve especially mid- and long-term action.
We thus provide answers to the unsettled question why the involvement
of the same neurobiological brain circuit, the DMN, has been documented
for baseline house-keeping functions and specific task performances.

In fact,
environmental stimuli especially important for humans are frequently of
social nature. This is probably unsurprising
given that a key property of the human species is
the complexity of their social systems
\citep{tomasello2009cultural}.
According to the ``social brain hypothesis'',
the human brain did not evolve to solve problems of the
physical environment, but to form and maintain increasingly complex
social systems to solve ecological problems by means of social relationships
\citep{whiten1988machiavellian}.
This may explain why social topics amount to roughly
two thirds of human everyday communication \citep{dunbar1997human}
and
why mind-wandering and dreams
are so rich in stories about people and
the complex relationships between them.
%
In line with this, the DMN has been proposed to reflect
continuous processing of social information as a
physiological baseline of human brain function
\citep{schilbach2008minds}. This view was later challenged by observing
analogues of the DMN in monkeys \citep{mantini2011default}
and rats \citep{lu2012rat}, species with
supposedly less advanced social-cognitive capacities.
Instead,
our formal account can parsimoneously incorporate
the dominance of social content in
human mental activity by reexpressing relevance to humans
in high value function estimates for social information
\citep{baker2009action}.
The DMN may thus exert control on the human agent to prioritize
appraisal of and action towards social contexts,
without completely neglecting environmental cues of the physical world.
In sum,
our account on the DMN directly implies
its previously proposed ``radar'' function
of monitoring the environment for salient information
in general and inherently incorporates the importance of social environmental cues
as an important special case, without excluding other topics
contemplated in continuous mental activity.

>>>>

\subsection{Putting everything together}
The DMN is today known to consistently increase in neural
  activity when humans engage in goal-directed behavior that are detached from
  current sensory environment \citep{kenet2003spontaneously, fiser2004small}
  and it was proposed to be situated at the top of the functional network hierarchies
  \citep{carhart2010default, margulies2016situating}.
  Its involvement in thinking about the past,
  the future, and hypothetical possibilities ties in with the implicit computation of
  action and state cascades as a function of what happened in the past.
  A policy represents the repertoire of possible actions
    on the world given a current state. It encodes the probabilites of
    choosing actions to be executed in a certain situation.
% Interactions between external and internal sources of
% information are probably commonplace during perception \citep{gershman2010learning}
The DMN may subserve
  constant exploration of possible future actions and their
  cumulative reward outcomes. Implicit computation of future choices
  provides an explanation for the
  evolutionary emergence and practical usefulness of mind-wandering
  in humans.    
\begin{figure}[!h]
  \includegraphics[width=.9\linewidth]{neurobiological_and_rl_overview_DMN.pdf}
  \caption{Neurobiological implementation of RL by the DMN...}
  \label{fig:rl_process_chart}
\end{figure}

@@

\subsection{Sentinel hypothesis}
Processing self-relevant information was perhaps the first
cognitive account that was proposed for the DMN \citep{gusnard2001medial}.
Since then,
many investigators have speculated that neural activity in the DMN
may reflect the brain's continuous tracking of
relevance in the environment
as an advantageous evolutionary adaptation \citep{randy2008, hahn2007cingulate}.
According to this cognitive account, the human brain's baseline realizes
a ``radar'' function to
detect subjectively salient cues and unexpected events in the environment.
Propositions of such a sentinel function have however seldom detailed
the mechanisms of
how attention and memory resources are exactly reallocated when
encountering a salient environmental stimulus.
In the present formal account,
promising action trajectories
are recursively explored by human agents. Conversely,
certain branches of candidate action trajectories
are detected to be less worthy to become mentally explored.
This mechanism, expressed by the Bellman equation,
directly implies stratified allocation of attention and working memory load
over relevant cues and events in the environment.
%
Further,
our account provides a parsimonious explanation for
the increasingly recognized observation
of the DMN implication
in certain goal-directed experimental tasks and
in task-unconstrained mind-wandering \citep{smith2009, bzdok2016formal}.
Both cognitive sets may entail DMN recruitment
if environmental relevance is processed, manipulated, and
used for predicting the future.
During tasks,
the policy and value estimates may be
updated to optimize especially short-term action.
At rest, these parameter updates may
improve especially mid- and long-term action.
The horizon is expressed in the $\gamma$ parameter in the formal account.
We thus provide answers for the unsettled question why the involvement
of the same neurobiological brain circuit (i.e., DMN) has been documented
for baseline house-keeping functions and specific task performances.

More specifically,
environmental stimuli especially important for humans are frequently of
social nature. This is probably unsurprising
given that a likely human-defining property is
the complexity of the social systems
\citep{tomasello2009cultural}.
According to the ``social brain hypothesis'',
the human brain has not primarily evolved to solve problems of the
physical environment, but to form and maintain increasingly complex
social systems to solve ecological problems via social relationships
\citep{whiten1988machiavellian}.
Indeed, social topics amount to roughly
two thirds of human everyday communication \citep{dunbar1997human},
while
mind-wandering at daytime and dreams during sleep
are rich in stories about people and
the complex relationships between them.
%
In line with this, the DMN has been proposed to reflect
continuous processing of social information as a
physiological baseline of human brain function
\citep{schilbach2008minds}. This view was later challenged by observing
analogues of the DMN in monkeys \citep{mantini2011default}
and rats \citep{lu2012rat}, both species with
supposedly less advanced social-cognitive capacities.
Instead,
the formal account can parsimoneously incorporate
the dominance of social content in
human mental activity by reexpressing relevance of information about humans
in high value function estimates
\citep{baker2009action, kampe2001psychology, krienen2010clan}.
The DMN may thus exert control on the human agent in a way that prioritizes
appraisal of and action towards social contexts,
without entirely neglecting environmental cues of the physical world.
In sum,
our account on the DMN directly implies
its previously proposed ``sentinel'' function
of monitoring the environment for salient information
in general and
inherently incorporates the importance of social environmental cues
as an important special case.
